Conjugate Gradient, Boosting, and Herding: \\ Connections and Application

### 2014-02-05 arXiv/introduction.tex
created new arXiv version and updated notation

In order to define herding, let $\sX$ be an observation space, $\phi : \sX \to \sB_{\Theta}$ a 
feature map to an inner product space, and $\bar\phi = \bE_{x \sim p}[\phi(x)]$ the empirical 
moments of some distribution. Then the herding algorithm is the following simple procedure to 
generate pseudosamples from $\sX$:
\[
x_{t} &\in \argmin_{x \in \sX} \ip{\theta_{t}}{\phi(x)} \\
\theta_{t} &= \theta_{t-1}  + \bar \phi - \phi(x_{t}).
\]
In this paper we generalize the herding algorithm to minimize $h(h) + R^{*}(u)$ via the following set of updates, which 
require only the convexity of $h$ and $R$:
\[
u_t &\in \argmin_{u} h(u) + \ip{\theta_t}{u} \\
\theta_{t+1} &\in \argmax_{\theta} \ip{\theta}{\hat{u}_t} - R(\theta),
\]
where $\hat{u}_t$ is a certain weighted average of $u_1,\ldots,u_t$. Our 
algorithm is described in more detail in Section~\ref{sec:algorithm}. This 
generalization has several advantages:
\begin{itemize}
\item It allows us to cast herding in the framework of mirror descent \cite{Beck:2003}, 
      which lets us better understand its convergence properties.
\item It allows us to obtain generalizations of herding for infinite 
      dimensional spaces that apply in cases where standard herding does not.
\item It makes the known connections between herding, subgradient, and 
      conditional gradient methods more clear. 
\end{itemize}
In particular, while it was known that herding was a form of conditional 
gradient descent \cite{Bach:2012a} and hence (via the relationship in \citet{Bach:2012b}) also 
a form of subgradient descent, our framework exposes all three algorithms 
as special cases. A full table outlining the relationships for various choices of 
$h$ and $R$ is given in Table~\ref{tab:connections}.

>>>>

In order to define herding, let $\sX$ be an observation space, $\phi : \sX \to \sF$ a 
feature map to an inner product feature space, and $\bar\phi = \bE_{x \sim p}[\phi(x)]$ the empirical 
moments of some distribution. Then the herding algorithm is the following simple procedure to 
generate pseudosamples from $\sX$:
\[
x_{t} &\in \argmin_{x \in \sX} \ip{\phi(x)}{y_{t-1}} \\
y_{t} &= y_{t-1}  + \bar \phi - \phi(x_{t}).
\]
In this paper we generalize the herding algorithm to minimize $\psi(x) + R^{*}(x)$ via the following set of updates, which 
require only the convexity of $\psi$ and $R$:
\[
x_t &\in \argmin_{x} \psi(x) + \ip{x}{y_{t-1}} \\
y_{t} &\in \argmax_{y} \ip{\bar{x}_t}{y} - R(y),
\]
where $\bar{x}_t$ is a certain weighted average of $x_1,\ldots,x_t$. Our 
algorithm is described in more detail in Section~\ref{sec:algorithm}. This 
generalization has several advantages:
\begin{itemize}
\item It allows us to cast herding in the framework of mirror descent \cite{Beck:2003}, 
      which lets us better understand its convergence properties.
\item It allows us to obtain generalizations of herding for infinite 
      dimensional spaces that apply in cases where standard herding does not.
\item It makes the known connections between herding, subgradient, and 
      conditional gradient methods more clear. 
\end{itemize}
In particular, while it was known that herding was a form of conditional 
gradient descent \cite{Bach:2012a} and hence (via the relationship in \citet{Bach:2012b}) also 
a form of subgradient descent, our framework exposes all three algorithms 
as special cases. A full table outlining the relationships for various choices of 
$\psi$ and $R$ is given in Table~\ref{tab:connections}.

### 2014-02-05 arXiv/lower-bounds.tex
created new arXiv version and updated notation

Although it is possible to achieve the fast $O(1/T^{2})$ convergence rate for finite-dimensional herding, \citet{Bach:2012a} note that without further assumptions herding has only a $O(1/T)$ convergence rate in the infinite-dimensional case. We now provide an example showing this bound is tight for {\em any} herding-like algorithm.  By ``herding-like,'' we mean algorithms that approximate the distribution with at most one additional sample at each step. We formalize this idea by defining a {\em sparse update method} to be any method that produces an estimator of $\bar \phi$ at iteration $t$ of the form 
\[
\theta_{t} = \sum_{s =1}^{t}w_{t} \phi(x_{t}),
\]
where $w_{t} \in \bR$. Sparse update methods include herding, $\alpha_{t}$-weighted herding, \cgd, and  \cgd with line search. These sparse methods -- which only operate on the extreme points of the simplex -- are of interest because they provide (weighted) ``samples'' matching the moments of the distribution of interest.

Our example is constructed as follows. Let the sample space be $\sX = \{ 1, 2, 3, \dots\}$. Define the infinite feature vector $\phi$ to be $\phi_{n}(x) = \bI(x = n)$, so $\sM$ is the unit $\ell^{1}$ ball. For target distribution $p \in \Delta_{\sX}$, we therefore have  that the mean vector is the vector of probabilities: $\bar\phi_{n} = p(n)$. Let $p(x) = c/\alpha(n)$, where $\alpha(n)$ is any superlinear function (i.e.~$\alpha \in \omega(n)$) and $c^{-1} = \sum_{n=1}^{\infty} \alpha(n) < \infty$. Without loss of generality we can assume that $\alpha$ is monotonically increasing.  We will consider the standard $\ell^{2}$ herding loss function $L(\theta) = \|\theta - \bar \phi\|_2^2$.

Since we are only considering sparse update methods, the $t$-th estimator will have at most $t$ non-zero entries, and so 
\[
\|\theta_{t}- \bar \phi\|_2^2 \ge c^{2}\sum_{n > t} 1/\alpha(n)^{2} = \Theta(1/\beta(t)) \label{eq:convergence-lb}
\]
for some (monotonically increasing) superlinear function $\beta$. But for any such $\beta$ we can choose some $\alpha$ to obtain \eqref{eq:convergence-lb}. In particular, let $1/\gamma(t) = \frac{\dee}{\dee t}\frac{1}{\beta(t)}$, where $\gamma(t) \in \omega(t^{2})$. Then we can choose $\alpha = \sqrt \gamma \in \omega(t)$ to satisfy \eqref{eq:convergence-lb}. Therefore, any sparse update method must have convergence rate no faster than $\Theta(1/\beta(t))$ for all superlinear $\beta$, and hence no faster than $\Theta(1/t)$.

>>>>

Although it is possible to achieve the fast $O(1/T^{2})$ convergence rate for finite-dimensional herding, \citet{Bach:2012a} note that without further assumptions herding has only a $O(1/T)$ convergence rate in the infinite-dimensional case. We now provide an example showing this bound is tight for {\em any} herding-like algorithm.  By ``herding-like,'' we mean algorithms that approximate the distribution with at most one additional sample at each step. We formalize this idea by defining a {\em sparse update method} to be any method that produces an estimator of $\bar \phi$ at iteration $t$ of the form 
\[
y_{t} = \sum_{s =1}^{t}w_{t} \phi(x_{t}),
\]
where $w_{t} \in \bR$. Sparse update methods include herding, $\alpha_{t}$-weighted herding, \cgd, and  \cgd with line search. These sparse methods -- which only operate on the extreme points of the simplex -- are of interest because they provide (weighted) ``samples'' matching the moments of the distribution of interest.

Our example is constructed as follows. Let the sample space be $\sX = \{ 1, 2, 3, \dots\}$. Define the infinite feature vector $\phi$ to be $\phi_{n}(x) = \bI(x = n)$, so $\sM$ is the unit $\ell^{1}$ ball. For target distribution $p \in \Delta_{\sX}$, we therefore have  that the mean vector is the vector of probabilities: $\bar\phi_{n} = p(n)$. Let $p(x) = c/\alpha(n)$, where $\alpha(n)$ is any superlinear function (i.e.~$\alpha \in \omega(n)$) and $c^{-1} = \sum_{n=1}^{\infty} \alpha(n) < \infty$. Without loss of generality we can assume that $\alpha$ is monotonically increasing.  We will consider the standard $\ell^{2}$ herding loss function $L(y) = \|y - \bar \phi\|_2^2$.

Since we are only considering sparse update methods, the $t$-th estimator will have at most $t$ non-zero entries, and so 
\[
\|y_{t}- \bar \phi\|_2^2 \ge c^{2}\sum_{n > t} 1/\alpha(n)^{2} = y(1/\beta(t)) \label{eq:convergence-lb}
\]
for some (monotonically increasing) superlinear function $\beta$. But for any such $\beta$ we can choose some $\alpha$ to obtain \eqref{eq:convergence-lb}. In particular, let $1/\gamma(t) = \frac{\dee}{\dee t}\frac{1}{\beta(t)}$, where $\gamma(t) \in \omega(t^{2})$. Then we can choose $\alpha = \sqrt \gamma \in \omega(t)$ to satisfy \eqref{eq:convergence-lb}. Therefore, any sparse update method must have convergence rate no faster than $y(1/\beta(t))$ for all superlinear $\beta$, and hence no faster than $y(1/t)$.

### 2014-02-05 arXiv/paper.tex
created new arXiv version and updated notation

\documentclass{article}
\usepackage{import, subfiles}
\usepackage{hyperref, icml2014}
\usepackage{tabularx}
\input{latex-defs.tex}

@@

\icmltitlerunning{Herding and Boosted Mirror Descent}

@@

\twocolumn[
\icmltitle{Herding and its Generalizations via Boosted Mirror Descent}

\icmlauthor{Jonathan Huggins \NA{add footnote on authorship}}{jhuggins@mit.edu}
\icmladdress{MIT CSAIL,
             77 Massachusetts Avenue, Cambridge, MA 02139 USA}
\icmlauthor{Jacob Steinhardt}{jsteinhardt@cs.stanford.edu}
\icmladdress{Stanford University,
             353 Serra Street, Stanford, CA 94305 USA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boosting, optimization, mirror descent, frank-wolfe, conjugate gradient descent, herding}

\vskip 0.3in
]

@@

\subfile{introduction.tex}
\subfile{boosted-mirror-descent.tex}
\subfile{herding-connection.tex}
\subfile{convergence-proofs.tex}
\subfile{lower-bounds.tex}
%\subfile{chen-proofs.tex}
\subfile{conclusion.tex}

@@

\end{document}

>>>>

\documentclass[reqno,oneside,a4paper]{amsart}
\usepackage{import, subfiles}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{chngpage}
\input{latex-defs.tex}

@@

@@

\title{Conjugate Gradient, Boosting, and Herding: \\ 
	   Connections and Application}

\thanks{The authors contributed equally to this work.}

\author[J.~H.~Huggins]{Jonathan H.~Huggins}
\address{Massachusetts Institute of Technology}
\urladdr{http://jhuggins.org/}
\email{jhuggins@mit.edu}

\author[J.~Steinhardt]{Jacob Steinhardt}
\address{Stanford University}
\urladdr{http://cs.stanford.edu/~jsteinhardt/}
\email{jsteinhardt@cs.stanford.edu}

\keywords{boosting, optimization, mirror descent, Frank-Wolfe, conjugate gradient descent, herding, graph sparsification}

@@

\maketitle

\subfile{introduction.tex}
\subfile{boosted-mirror-descent.tex}
\subfile{herding-connection.tex}
\subfile{convergence-proofs.tex}
\subfile{lower-bounds.tex}
\subfile{chen-proofs.tex}
\subfile{conclusion.tex}

@@

\end{document} \maketitle
